---
title: "Class 11: AlphaFold"
author: "Melissa (PID: A19149673)"
format: pdf
toc: TRUE
---

## Custom Analysis of Resulting Models

```{r}
library(bio3d)
pdb <- read.pdb("hivp_dimer_0d832/hivp_dimer_0d832_unrelaxed_rank_001_alphafold2_ptm_model_1_seed_000.pdb")
pdb
```

Make a vector of input PDB file names that we can read into R

```{r}
pdb_files <- list.files("hivp_dimer_0d832/", 
           pattern =".pdb",
           full.names = TRUE)
basename(pdb_files)
```
```{r}
# Read all data from Models 
#  and superpose/fit coords
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")
pdbs
```
```{r}
rd <- rmsd(pdbs, fit=T)
range(rd)
```

```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```
The RMSD heatmap compares structural similarity between the five AlphaFold models. Additionally, the lower RMSD values indicate more similar structures while models with higher RMSD are more structurally different from the others.

```{r}
pdb <- read.pdb("1hsg")
```

```{r}
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```
The pLDDT plot shows AlphaFold confidence across residue positions. Moreover, the higher pLDDT values indicate higher confidence, while lower values suggest less reliable regions of the model.

```{r}
core <- core.find(pdbs)
```

```{r}
core.inds <- print(core, vol=0.5)
```

```{r}
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")

```

```{r}
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```

The RMSF plot shows how much each residue varies across the fitted models as the higher RMSF values indicate more flexible or variable regions.

## Predicted alignment error for domains

```{r}
library(jsonlite)
# Listing of all PAE JSON files
results_dir <- "hivp_dimer_0d832/"
pae_files <- list.files(path=results_dir,
                        pattern=".*model.*\\.json",
                        full.names = TRUE)
```

```{r}
pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae1)
```

```{r}
# Per-residue pLDDT scores 
#  same as B-factor of PDB..
head(pae1$plddt)
```

```{r}
pae1$max_pae
pae5$max_pae
```

The maximum PAE values compare prediction uncertainty between models. A lower max PAE indicates a more reliable model. Hence, the model 1 is more reliable than the model 5.

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```

```{r}
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```
The PAE plots show predicted alignment error between residue pairs. The first Model 1 shows lower overall error than model 5. This suggests a more confident domain arrangement.

## Residue conservation from alignment file

```{r}
aln_file <- list.files(path=results_dir,
                       pattern=".a3m$",
                        full.names = TRUE)
aln_file
```

```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)
```

```{r}
dim(aln$ali)

```

```{r}
sim <- conserv(aln)
plotb3(sim[1:99], sse=trim.pdb(pdb, chain="A"),
       ylab="Conservation Score")
```
The conservation plot shows which residues are most conserved across the sequence alignment.

```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```

```{r}
m1.pdb <- read.pdb(pdb_files[1])

# uses conservation scores for all residues in the model
sim2 <- rep(sim[1:99], 2)

# map residue conservation scores onto every atom
occ <- sim2[match(m1.pdb$atom$resno, sort(unique(m1.pdb$atom$resno)))]

# write new PDB file for Mol Star
write.pdb(m1.pdb, o = occ, file = "m1_conserv.pdb")
```

The original code produced a length mismatch error, so it was adjusted to map conservation scores to residue numbers.
