---
title: "Class 12 Genome Informatics"
author: 'Melissa (PID: A19149673)'
date: "2026-05-08"
output: pdf_document
---

## Section 4: Population Scale Analysis

> Q13. Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. 

```{r}
expr <- read.table("https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
tapply(expr$exp, expr$geno, median)
```

The sample sizes are 108 for A/A, 233 for A/G and 121 for G/G. The median expression levels are about 31.25 for A/A, 25.06 for A/G and 20.07 for G/G.

> Q14.Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)
```

Let's make a boxplot

```{r}
ggplot(expr, aes(geno, exp, fill=geno)) + 
  geom_boxplot(notch=TRUE, outlier.shape = NA) + 
  geom_jitter(shape=16, position=position_jitter(0.2), alpha=0.4)
```

The ORMDL3 expression is highest in A/A, intermediate in A/G and lowest in G/G. Hence, this suggests the rs8067378 genotype is associated with ORMDL3 expression, with the G/G genotype showing lower expression.
