---
title: "Class 13: RNASeq analysis with DESeq"
author: "Melissa (PID: A19149673)"
format: pdf
toc: true
---
## Background

Today we're going to do an RNASeq analysis with DESeq.

## Bioconductor Setup

```{r}
library(DESeq2)
```

## Data Import

Let's read the `count` data and `metadata` about this experiment setup from the supplied CSV files:

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names = 1)
metadata <- read.csv("airway_metadata.csv")
```

have a wee peak:
```{r}
head(counts)
```

and the metadata that tells us what is actually in the columns of our `counts` object:
```{r}
metadata
```

> Q1. How many genes are in this dataset?

There are `r nrow(counts)` , or 38694 genes in this dataset

> Q2. How many `control` cell lines do we have?

```{r}
sum( metadata$dex == "control" )
```
```{r}
table(metadata$dex)
```
There are 4 `control` cell lines.

```{r}
colnames(counts) == metadata$id
```

```{r}
metadata
```

## Toy differential gene expression

- find the "control" columns in our `counts` object
- Extract just the "control" column values for all genes
- Calculate the average value per gene in these "control" columns

```{r}
control.inds <- metadata$dex == "control"
control.counts <- counts[,control.inds]
control.mean <- rowMeans(control.counts)
```

```{r}
control.mean
```

Now do the same for the "treated" columns. Then make a plot of `control.mean` vs `treated.mean`

```{r}
treated.inds <- metadata$dex == "treated"
treated.counts <- counts[, treated.inds]
treated.mean <- rowMeans(treated.counts)
head(treated.mean)
```

```{r}
meancounts <- data.frame(control.mean, treated.mean)
head(meancounts)
```


> Q3. How would you make the above code in either approach more robust? Is there a function that could help here?


```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts ) 
head(control.mean)
```

Using rowSums() makes the code more robust. And if additional control samples are added later, the code will still calculate the correct average automatically.

> Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)

```{r}
treated <- metadata[metadata$dex == "treated", ]
treated.counts <- counts[, treated$id]
treated.mean <- rowMeans(treated.counts)
treated.mean
```


```{r}
meancounts <- data.frame(control.mean, treated.mean)
```

> Q5. (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples. Your plot should look something like the following.

```{r}
plot(meancounts$control.mean, meancounts$treated.mean)
```

> Q5. (b).You could also use the ggplot2 package to make this figure producing the plot below. What geom_?() function would you use for this plot? 

```{r}
library(ggplot2)

ggplot(meancounts, aes(x = control.mean, y = treated.mean)) +
  geom_point()
```

We would use geom_point() function for this plot.

> Q6. Try plotting both axes on a log scale. What is the argument to plot() that allows you to do this? 

Our count data is highly skewed and when we see a pattern like this plot it SCREAMS log transform me!

```{r}
plot(meancounts$control.mean,
     meancounts$treated.mean,
     log="xy",
     xlab="log Control",
     ylab="log Treated")
```

The argument that allows log scaling is, log = "xy".

We most often use log2 transform for this kind of data in bioinformatics because it makes my brain hurt less.

```{r}
log2(20/40)
```

```{r}
#Treated / Control

log2(20/20)
log2(40/20)
log2(10/20)
log(80/20)
```

We call this little fraction the "log2 fold change" as it tells us how much more or less gene expression we have in units of doubling, etc.

Let's calculate the log2 fold change for our `treated.mean` and `control.mean` counts and call this `log2fc`.

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

A common "rule of thumb" threshold for calling a gene "up regulated" or "down regulated" us a log2 fold-change value of +2 or -2 (or greater)

> Q7. What is the purpose of the arr.ind argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?

```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

The arr.ind=TRUE causes which() function to return both row and column positions where true values occur. We take the first column since it contains the row numbers genes with zero counts. And we use unique() function so the same gene is not counted multiple times if it contains zeros in multiple samples.

> Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2 fc level? 

```{r}
up.ind <- mycounts$log2fc > 2

sum(up.ind)

paste("Up:", sum(up.ind))
```

> Q9. Using the down.ind vector above can you determine how many down regulated genes we have at the greater than 2 fc level? 

```{r}
down.ind <- mycounts$log2fc < (-2)

sum(down.ind)

paste("Down:", sum(down.ind))
```

> Q10. Do you trust these results? Why or why not?

No, I don't fully trust these results since they are only based on fold-change and do not account for statistical significance. Additionally, large fold changes can occur by chance and we have not yet calculated and adjusted p-values. Overall, DESeq2 performs this analysis more properly using statistical models for RNA-seq count data.

## Setting up for DESeq

Let's do this analyze properly and not forget about the significance of the differences.

For this we will use **DESeq2** package

```{r}
library(DESeq2)
citation("DESeq2")
```

To run a DESeq analysis we need at least two inputs: 
- `countData` i.e. are gene counts across different expriments

- `colData` i.e. our metadata about those count columns

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```

```{r}
dds <- DESeq(dds)
res <- results (dds)
head(res)
```

## Principal Component Analysis (PCA)

```{r}
vsd <- vst(dds, blind = FALSE)
plotPCA(vsd, intgroup = c("dex"))

```

```{r}
pcaData <- plotPCA(vsd,
                   intgroup=c("dex"),
                   returnData=TRUE)
head(pcaData)
```

```{r}
# calculating percent variance per PC for the plot axis labels
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

```{r}
ggplot(pcaData) +
  aes(x=PC1,
      y=PC2,
      color=dex) +
  geom_point(size=3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  theme_bw()
```
## DESeq Analysis

```{r}
results(dds)
```

```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
res
```

```{r}
summary(res)
```

```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
```

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
columns(org.Hs.eg.db)
```

```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), # our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="SYMBOL",          # The new format we want to add
                     multiVals="first")
head(res)
```

> Q11. Run the mapIds() function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called res$entrez, res$uniprot and res$genename.

```{r}
res$entrez <- mapIds(org.Hs.eg.db,
                     keys = row.names(res),
                     column = "ENTREZID",
                     keytype = "ENSEMBL",
                     multiVals = "first")

res$uniprot <- mapIds(org.Hs.eg.db,
                      keys = row.names(res),
                      column = "UNIPROT",
                      keytype = "ENSEMBL",
                      multiVals = "first")

res$genename <- mapIds(org.Hs.eg.db,
                       keys = row.names(res),
                       column = "GENENAME",
                       keytype = "ENSEMBL",
                       multiVals = "first")

head(res)
```


```{r}
ord <- order(res$padj)

head(res[ord,])
```

```{r}
write.csv(res[ord,], "deseq_results.csv")
```

## Volcano Plot

```{r}
plot(res$log2FoldChange,
     -log(res$padj),
     xlab="Log2(FoldChange)",
     ylab="-Log(P-value)")
```

This is a ubiquitois and common visualization for this type of data that puts the log2 foldchange and the adjusted p-value together in one plot that people can get insight for what is going on in the whole dataset results.

The plot is very useful because we don't care about genes with high p-values  - log2foldchange


```{r}
plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.05), col="darkgray", lty=2)
```
> Q. Add anotations to this volcano plot including the log2 fold-change thresholds of +2 and -2 and the p-value threshold of 0.05. Also color up just the genes that meet both these thresholds. These are the ones we will focus on next day!

```{r}
# setup our custom point color vector 
mycols <- rep("gray", nrow(res))

# color genes with large fold changes red
mycols[abs(res$log2FoldChange) > 2] <- "red"

# color genes meeting BOTH thresholds blue
inds <- (res$padj < 0.05) & 
        (abs(res$log2FoldChange) > 2)

mycols[inds] <- "blue"

# volcano plot with custom colors
plot(res$log2FoldChange,
     -log(res$padj),
     col = mycols,
     ylab = "-Log(P-value)",
     xlab = "Log2(FoldChange)")

# cutoff lines
abline(v = c(-2, 2), col = "gray", lty = 2)
abline(h = -log(0.05), col = "gray", lty = 2)
```

The volcano plot includes annotations for log2 fold-change thresholds of +2 and -2 and a p-value threshold of 0.05. Moreover, the genes meeting both thresholds are highlighted in blue.

## Save our results to date

```{r}
write.csv(res, file="myresults.csv")
```

```{r}
library(EnhancedVolcano)
```

```{r}
x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```

## Pathway Analysis

```{r}
library(pathview)
library(gage)
library(gageData)

data(kegg.sets.hs)

# the first 2 pathways in this kegg set for humans
head(kegg.sets.hs, 2)
```

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
# get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

```{r}
attributes(keggres)
```

```{r}
head(keggres$less, 3)
```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa05310")
```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa05310", kegg.native=FALSE)
```

```{r}
head(keggres$less, 2)
```

> Q12.  Can you do the same procedure as above to plot the pathview figures for the top 2 down-reguled pathways?

```{r}
pathview(gene.data = foldchanges, pathway.id = "hsa05332", kegg.native = FALSE)
pathview(gene.data = foldchanges, pathway.id = "hsa04940", kegg.native = FALSE)
```

Question 12 plots the pathview figures for the top two down-regulated KEGG pathways from keggres$less. This includes Graft versus host disease and Type I diabetes mellitus.