---
title: "HW Class 6 (R Functions)"
author: 'Melissa (PID: A19149673)'
output: pdf_document
---

## Original code
```{r}
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
# there is a small error in original code 
#where s1 is used instead of s3 below
s3.chainA <- trim.pdb(s1, chain="A", elety="CA")
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")

```

## Homework Class 6

> Q6: How would you generalize the original code above to work with any set of input protein structures?

```{r}
library(bio3d)

# the plot_protein() function reads a protein structure from the PDB database 
# and extracts B-factor values and generates a plot for visualization of these 
# values
# the inputs include: pdb_id, chain, and atom_type.
# pdb_id is a character that specifies the PDB ID, the chain is the protein chain
# additionally, the atom_type is the atom type to extract
# the outputs include a plot of B-factors and residue position as well as 
# a numeric vector of the B-factor values

plot_protein <- function(pdb_id, chain = "A", atom_type = "CA") {
  
  # the PDB file is read 
  pdb <- read.pdb(pdb_id)
  
  # the trim function is used to trim to chain A and alpha carbons
  pdb_trim <- trim.pdb(pdb, chain = chain, elety = atom_type)
  
  # the b-factor values is extracted from the trimmed structure
  b_factors <- pdb_trim$atom$b
  
  # b-factors are plotted across residues
  # the typ ="l" creates a line plot
  # the sse adds secondary structure such as helix and sheet regions
  plotb3(b_factors, sse = pdb_trim, typ = "l", ylab = "Bfactor")
  
  # return values of the b-factor values 
  return(b_factors)
}
```

## Example output run
```{r}
plot_protein("4AKE")
plot_protein("1AKE")
plot_protein("1E4Y")
```

The plot_protein() function takes a PDB ID such as "4AKE" for the input and the optional arguments for the chain and atom type. The function serves to read a protein structure using the read.pdb() function which trims it to chain A alpha carbons using the trim function, trim.pdb(). This essentially extracts the B-factor values and plots to visualize the protein. Additionally, the function produces an output of a B-factor plot and returns a numeric vector of B-factor values.
