---
title: "Class 7 Machine Learning 1"
author: "Melissa (PID: A19149673)"
format: pdf
toc: true
---

## Background

Today we will explore some core machine learning methods that are very popular in bioinformatics. These include **clustering** and **dimensionality reduction**.

## K-means clustering

The main function in "base" R for K-means clustering is called `kmeans()`

Before we go too deep let's make up some "simple" data that we can cluster and know if we are getting agood answer or not. To do this we can use `rnorm()`
function:

```{r}
hist( rnorm (10000, mean = 3))
```
```{r}
x <- c( rnorm(30, -3),rnorm(30, +3))
z <- cbind(x=x, y = rev(x))
plot(z)
```

```{r}
p <- 1:5
rev(p)
```
Now we can run `kmeans()` on this input `z` and see what the results look like.

```{r}
km <- kmeans(z, centers = 2)
km
```
```{r}
attributes(km)
```

> Q. How many points are in each cluster?

```{r}
km$size
```

> Q. What "component of your result object details cluster assignment/membership"?

```{r}
km$cluster
```

> Q. What "component of your result object details cluster center"?

```{r}
km$centers
```

> Q. Plot `z` colored by the kmeans cluster assignment and add cluster centers as blue points.

```{r}
plot(z, col=km$cluster)
points(km$centers, col = "blue", pch = 15)
```
> Q. Run a K-means clustering and plot the results asking for 4 clusters? (k=4)?

```{r}
km4 <- kmeans(z, centers = 4)
plot (z, col=km4$cluster)
points(km4$centers, col ="black", pch = 15)
```
> **N.B.** You need to tell K-means the number of clusters (i.e. set `centers=2`)!!

One approach is to try different values for `centers` and then pick the best...

```{r}
ans <- NULL
for(i in 1:10) {
km <- kmeans(z, centers=i)
ans <- c(ans, km$tot.withinss)
}

plot(ans, typ="o", xlab = "Number of clusters",
     ylab = "Total Sum of Squares Distance")
```

## Hierarchial Clustering

The main function in "base" R for Hierarchial Clustering is called `hclust()`

This function does not take your "raw" data for clustering. You must first build a "distance matrix" from your data and pass this as input to `hclust()`

```{r}
d <- dist(z)
hc <- hclust(d)
hc
```

There is a bespoke `plot()` method for `hclust()` result objects.
```{r}
plot(hc)
abline(h = 3, col = "red")
```

Once we have our `hclust` object (our "tree" of "cluster dendrogram") we can *"cut"* the tree to reveal the clustering pattern.

```{r}
cutree(hc, k=4)
```

> Q. Make a plot of `z` with your hclust results (i.e. colored by cluster membership)

```{r}
grps <- cutree(hc, k = 4)
plot(z, col= grps, pch =16)
```

## Principal Component Analysis (PCA)

PCA is a dimensionallity of reduction method that is popular for revealing patterns in complex datasets.

## Analysis of UK food data

Let's look at some data on the eating habits of folks from the UK to see if there are patterns and trends that have some regions being distinct from others.

## Data Import

The data is made available in CSV format so we can use `read.csv()` function for import to R:

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

## Tidy the data

Fix anything that went wrong with the data import.

## Exporatory analysis

Make some plots to help make sense of obvious trends...

## PCA

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

## Complete the following code to find out how many rows and columns are in x?

```{r}
dim(x)
```
##Preview the first 6 rows

```{r}
head(x)
```

```{r}
rownames(x) <- x[,1] # This keeps only column 1
x <- x[,-1] #This keeps everything except column 1
head(x)
```
```{r}
dim(x)
```
There are 17 rows and 4 columns after fixing the row names.

```{r}
x <- read.csv(url, row.names=1)
head(x)
```
> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

Setting row names during import is more robust since it avoids extra steps and prevents error later.

> Q3. Changing what optional argument in the above barplot() function results in the following plot?

```{r}
cols <-rainbow(nrow(x))
barplot(as.matrix(x), col=cols)
```
```{r}
dim(x)
```

```{r}
head(x)
```
```{r}
library(tidyr)

x_long <- x |> 
          tibble::rownames_to_column("Food") |> 
          pivot_longer(cols = -Food, 
                       names_to = "Country", 
                       values_to = "Consumption")

dim(x_long)
```
```{r}
library(ggplot2)

ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "dodge") +
  theme_bw()
```

The change comes from changing the argument that controls how the bars are arranged. Also determines whether the barplot shows separate bars for each category as well as combined stack bars.

> Q4: Changing what optional argument in the above ggplot() code results in a stacked barplot figure?

```{r}
ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "stack") +
  theme_bw()
```

Switching the argument changes the plot from comparing categories separately to show their combined totals within each group.

> Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)

```

```{r}
library(pheatmap)
pheatmap(as.matrix(x))
```

The pairs() plot shows all pairwise comparison between the countries. This helps visually show how similar and different countries are based on their food consumption across the categories.

> Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

> **Key-point**: Even relatively small datasets can prove challenging to interpret.

## PCA to the rescue

```{r}
t(x)
```

The main function in "base" R for PCA is called `prompt()`. This function wants the observations to be rows and the variables to be columns.

```{r}
pca <- prcomp( t(x) )
summary(pca)
```
```{r}
attributes(pca)
```
The main result figure from this analysis is called "PC score plot" or "ordination plot" "PC olot" or PC1 vs PC2 plot".

This plot show samples (in this case countries) relate to each other along our new PC axis.

This is our new "reduced-dimensional space". In this case2 dimensions, PC1 and PC2, that capture most of the variance to the original 17 dimensional data-set.

```{r}
pca$x
```
England, Wales and Scotland cluster together to show similar diets while Northern Ireland is pretty distinct. However, the specific food differences require further analysis such as the PCA loading's to be able to identify.

> Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

> Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

```{r}
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

ggplot(df) +
  aes(x = PC1, y = PC2, label = Country, color = Country) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  scale_color_manual(values = c(
    "England" = "orange",
    "Wales" = "red",
    "Scotland" = "blue",
    "N.Ireland" = "darkgreen"
  )) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
```

```{r}
z <- summary(pca)
z$importance
```

```{r}
# Create scree plot with ggplot
variance_df <- data.frame(
  PC = factor(paste0("PC", 1:length(v)), levels = paste0("PC", 1:length(v))),
  Variance = v
)

ggplot(variance_df) +
  aes(x = PC, y = Variance) +
  geom_col(fill = "steelblue") +
  xlab("Principal Component") +
  ylab("Percent Variation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0))
```

```{r}
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```
> Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

```{r}
ggplot(pca$rotation) +
  aes(x = PC2, 
      y = reorder(rownames(pca$rotation), PC2)) +
  geom_col(fill = "steelblue") +
  xlab("PC2 Loading Score") +
  ylab("") +
  theme_bw()
```

The food groups that feature most prominently in PC2 are those with the largest positive and negative loading scores such as fish and fats_and_oils. These variables contribute strongly to the variation captured by PC2. PC2 also represents a secondary dietary pattern, as this distinguishes countries based on differences in specific types of food consumption rather than overall intake. Moreover, PC1 explains about 67% of the variance while PC2 explains about 29%.
